---
title: "Severe weather events" 
subtitle: "Their impact on public health and the economy."
author: Adam Blanch
date: 28th April 2020
output:
  github_document:
    html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

## Synopsis

The basic goal of this report is to explore the NOAA Storm Database and determine:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

## Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.


## Data Processing

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

* [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) [47Mb]

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

* National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
* National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

For a expense comparison the [CPI data](https://www.federalreserve.gov/datadownload/Output.aspx?rel=G17&series=865924caea53dd4e8edcf7cb134f80db&lastobs=&from=01/01/1950&to=12/31/2011&filetype=csv&label=include&layout=seriescolumn) for the period Jan 1950 to Dec 2011 is available to download.

### Assumptions

Processing the data has a few assumtpions:
1. The PROPDMGEXP and CROPDMGEXP numeric values represent powers of 10. i.e "3"would be equivalent to "k" or 1000 times.
2. Some of the large events cross multiple states and counties, the same damge figure is duplicated. For the purpose of this analysis I have allowed the values to be duplicated but further analysis may be required to determine the validity and impact on study results.

### Download and Read Data

```{r DataProcessing}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
zipFile <- "repdata_data_StormData.bzip2"
dataFile <- "stormdata.csv"
#this flag is TRUE if you want to store the csv on the local directory
#             FALSE if you want to extract the csv from the zip each time
storeCSV <- TRUE
#Download the data
if(!file.exists(zipFile))
{
  download.file(url,destfile=zipFile,method="curl")
}
#Extract the csv from the zip file
if(!file.exists(dataFile) && storeCSV)
{
  stormdata <- read.csv(zipFile)
  if(storeCSV)
    write.csv(stormdat,file = dataFile)
} else {
  stormdata <- read.csv(dataFile)
}

# only going to access some of the information so to save memory, just select those
# columns. NB: STATE and REMARKS are not used but help provide detail to check validity
# of assumtions
stormdata <- stormdata[,c("BGN_DATE","STATE","EVTYPE","FATALITIES","INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP","REMARKS")]

download.file("https://www.federalreserve.gov/datadownload/Output.aspx?rel=G17&series=865924caea53dd4e8edcf7cb134f80db&lastobs=&from=01/01/1950&to=12/31/2011&filetype=csv&label=include&layout=seriescolumn",destfile="cpi.csv",method="curl")
cpiData <- read.csv("cpi.csv",header=TRUE,skip=5)
baseCPI <- c(CPI=cpiData[cpiData$Time.Period=="2011-12","IP.B50001.S"],date="2011-12")
```

### Tidy Data

The National Climatic Data Center Storm Events lists 48 different event types but exploring the data fromm the csv file has `r length(levels(stormdata$EVTYPE))` different event types.  This includes typos, abbreviations and alternate names.  So the first step is to group the data by the 48 different event types using the [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) to make valid reclassifcations.  The code for this is included in [tidy_stormdata.R](https://github.com/AJB276/ReproducibleResearch)

``` {r TidyData}
source("tidy_stormdata.R")
```
### Summary

``` {r Summary}
sum(!complete.cases(stormdata$EVTYPENEW))
stormdata <- stormdata[complete.cases(stormdata$EVTYPENEW),]
stormdata$EVTYPENEW <- factor(stormdata$EVTYPENEW,exclude = NA)
stormdata$BGN_DATE <- as.POSIXct(stormdata$BGN_DATE,format="%m/%d/%Y %H:%M:%OS")

eventsummary <- stormdata %>% group_by(EVTYPENEW) %>% summarise(total_fatalities = sum(FATALITIES), mean_fatalities = mean(FATALITIES), count = n())

propsummary <- stormdata %>% group_by(EVTYPENEW,PROPDMGEXP) %>% summarise(total_propdmg = sum(PROPDMG), mean_propdmg = mean(PROPDMG), count = n())
propsummary[order(propsummary$count,decreasing = TRUE),]

damageValues <- function(dmg,dateStr,dmgStr,dmgEXPStr)
{
  dmg$year <- year(dmg[,dateStr])
  dmg$month <- month(dmg[,dateStr])
  dmg$CPI <- cpiData[(dmg$year-1950)*12+dmg$month,"IP.B50001.S"]
  suppressWarnings(dmg$power <- as.numeric(as.character(dmg[,dmgEXPStr])))
  dmg[is.na(dmg$power),]$power <- 0
  if(nrow(dmg[toupper(dmg[,dmgEXPStr])=="H",])) dmg[toupper(dmg[,dmgEXPStr])=="H",]$power <- 2
  dmg[toupper(dmg[,dmgEXPStr])=="K",]$power <- 3
  dmg[toupper(dmg[,dmgEXPStr])=="M",]$power <- 6
  dmg[toupper(dmg[,dmgEXPStr])=="B",]$power <- 9
  dmg$TOTDMG <- dmg[,dmgStr] * 10^dmg$power
  cpiBase <- as.numeric(baseCPI["CPI"])
  dmg$EQUIVTOTDMG <- dmg$TOTDMG*(cpiBase/dmg$CPI)
  dmg
}

propDMG <- stormdata[stormdata$PROPDMG > 0,c("EVTYPENEW","BGN_DATE","PROPDMG","PROPDMGEXP")]
propDMG <- damageValues(propDMG,"BGN_DATE","PROPDMG","PROPDMGEXP")

cropDMG <- stormdata[stormdata$CROPDMG > 0,c("EVTYPENEW","BGN_DATE","CROPDMG","CROPDMGEXP")]
cropDMG <- damageValues(cropDMG,"BGN_DATE","CROPDMG","CROPDMGEXP")
```

## Results

``` {r Results}
```

## Conclusion

Areas that arose from analysis of the data for further investigation or clarifcation was the detail on the property and crop damage.  An example of an event or events that have a large impact potential quality of the data is April 11, 1965 Palm Sunday Tornados.