---
title: "Severe weather events" 
subtitle: "Their impact on public health and the economy."
author: Adam Blanch
date: 28th April 2020
output:
  github_document:
    html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

## Synopsis

The basic goal of this report is to explore the NOAA Storm Database and determine:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

## Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.


## Data Processing

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

* [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) [47Mb]

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

* National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
* National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

For a expense comparison the [CPI data](https://www.federalreserve.gov/datadownload/Output.aspx?rel=G17&series=865924caea53dd4e8edcf7cb134f80db&lastobs=&from=01/01/1950&to=12/31/2011&filetype=csv&label=include&layout=seriescolumn) for the period Jan 1950 to Dec 2011 is available to download.

### Assumptions

Processing the data has a few assumtpions:

1. The PROPDMGEXP and CROPDMGEXP that contain numeric values represent powers of 10. i.e "3" would be equivalent to "k" or 1000 times.

2. Some of the large events cross multiple states and counties, the same damage figure is duplicated. For the purpose of this analysis I have allowed the values to be duplicated but further analysis may be required to determine the validity and impact on study results.

### Download and Read Data

```{r DataProcessing}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
zipFile <- "repdata_data_StormData.bzip2"
dataFile <- "stormdata.csv"
#this flag is TRUE if you want to store the csv on the local directory
#             FALSE if you want to extract the csv from the zip each time
storeCSV <- TRUE
#Download the data
if(!file.exists(zipFile))
{
  download.file(url,destfile=zipFile,method="curl")
}
#Extract the csv from the zip file
if(!file.exists(dataFile) && storeCSV)
{
  stormdata <- read.csv(zipFile)
  if(storeCSV)
    write.csv(stormdat,file = dataFile)
} else {
  stormdata <- read.csv(dataFile)
}

# only going to access some of the information so to save memory, just select those
# columns. NB: STATE and REMARKS are not used but help provide detail to check validity
# of assumtions
stormdata <- stormdata[,c("BGN_DATE","STATE","EVTYPE","FATALITIES","INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP","REMARKS")]

download.file("https://www.federalreserve.gov/datadownload/Output.aspx?rel=G17&series=865924caea53dd4e8edcf7cb134f80db&lastobs=&from=01/01/1950&to=12/31/2011&filetype=csv&label=include&layout=seriescolumn",destfile="cpi.csv",method="curl")
cpiData <- read.csv("cpi.csv",header=TRUE,skip=5)
baseCPI <- c(CPI=cpiData[cpiData$Time.Period=="2011-12","IP.B50001.S"],date="2011-12")
```

### Tidy Data

The National Climatic Data Center Storm Events lists 48 different event types but exploring the data fromm the csv file has `r length(levels(stormdata$EVTYPE))` different event types.  This includes typos, abbreviations and alternate names.  So the first step is to group the data by the 48 different event types using the [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) to make valid reclassifcations.  The code for this is included in [tidy_stormdata.R](http://github.com/AJB276/ReproducibleResearch)

``` {r TidyData}
# Code to tidy data into 48 different events is not included in report
source("tidy_stormdata.R")
```

``` {r CalculateDamage}
damageValues <- function(dmg,dateStr,dmgStr,dmgEXPStr)
{
  dmg$year <- year(dmg[,dateStr])
  dmg$month <- month(dmg[,dateStr])
  dmg$CPI <- cpiData[(dmg$year-1950)*12+dmg$month,"IP.B50001.S"]
  suppressWarnings(dmg$power <- as.numeric(as.character(dmg[,dmgEXPStr])))
  dmg[is.na(dmg$power),]$power <- 0
  if(nrow(dmg[toupper(dmg[,dmgEXPStr])=="H",])) dmg[toupper(dmg[,dmgEXPStr])=="H",]$power <- 2
  dmg[toupper(dmg[,dmgEXPStr])=="K",]$power <- 3
  dmg[toupper(dmg[,dmgEXPStr])=="M",]$power <- 6
  dmg[toupper(dmg[,dmgEXPStr])=="B",]$power <- 9
  dmg$TOTDMG <- dmg[,dmgStr] * 10^dmg$power
  cpiBase <- as.numeric(baseCPI["CPI"])
  dmg$EQUIVTOTDMG <- dmg$TOTDMG*(cpiBase/dmg$CPI)
  dmg
}
```

### Summary

``` {r Summary}
#sum(!complete.cases(stormdata$EVTYPENEW))
stormdata <- stormdata[complete.cases(stormdata$EVTYPENEW),]
stormdata$EVTYPENEW <- factor(stormdata$EVTYPENEW)
stormdata$BGN_DATE <- as.POSIXct(stormdata$BGN_DATE,format="%m/%d/%Y %H:%M:%OS")

eventsummary <- stormdata %>% group_by(EVTYPENEW) %>% summarise(total_fatalities = sum(FATALITIES), mean_fatalities = mean(FATALITIES), total_injuries = sum(INJURIES), mean_injuries = mean(INJURIES), count = n())

propDMG <- stormdata[stormdata$PROPDMG > 0,c("EVTYPENEW","BGN_DATE","PROPDMG","PROPDMGEXP")]
propDMG <- damageValues(propDMG,"BGN_DATE","PROPDMG","PROPDMGEXP")
propDMGsummary <- propDMG %>% group_by(EVTYPENEW) %>% summarise(total_damage = sum(TOTDMG), total_equivdamage = sum(EQUIVTOTDMG), count = n())

cropDMG <- stormdata[stormdata$CROPDMG > 0,c("EVTYPENEW","BGN_DATE","CROPDMG","CROPDMGEXP")]
cropDMG <- damageValues(cropDMG,"BGN_DATE","CROPDMG","CROPDMGEXP")
cropDMGsummary <- cropDMG %>% group_by(EVTYPENEW) %>% summarise(total_damage = sum(TOTDMG), total_equivdamage = sum(EQUIVTOTDMG), count = n())
```

## Results

### Across the United States, which types of events are most harmful with respect to population health?

The types of events that are most harmful with respect to population health have been plotted below:

``` {r HealthResults}
# events with highest fatalities
fatalitiesAll <- arrange(eventsummary,desc(total_fatalities),desc(total_injuries))
# events with highest injuries
injuriesAll <- arrange(eventsummary,desc(total_injuries),desc(total_fatalities))
par(mfrow = c(1, 2), mar = c(15, 4, 4, 2), mgp = c(3, 1, 0), cex = 0.7)
fatalities <- fatalitiesAll[1:10,]
injuries <- injuriesAll[1:10,]
barplot(fatalities$total_fatalities, las = 3, names.arg = fatalities$EVTYPENEW, main = "Events with Highest Fatalities", ylab = "Number of fatalities", col = "red")
barplot(injuries$total_injuries, las = 3, names.arg = injuries$EVTYPENEW, main = "Events with Highest Injuries", ylab = "Number of injuries", col = "yellow")
```

**Figure 1: Events with Highest Fatalities and Injuries**

### Across the United States, which types of events have the greatest economic consequences?

The types of events thathave the greatest economic consequences have been plotted below:

``` {r EconomicResults}
# events with highest property damage
propDMGorder <- arrange(propDMGsummary,desc(total_equivdamage))
# events with highest crop damage
cropDMGorder <- arrange(cropDMGsummary,desc(total_equivdamage))
par(mfrow = c(1, 2), mar = c(15, 4, 4, 2), mgp = c(3, 1, 0), cex = 0.7)
propDMGtop <- propDMGorder[1:10,]
cropDMGtop <- cropDMGorder[1:10,]
barplot(propDMGtop$total_equivdamage/(10^9), las = 3, names.arg = propDMGtop$EVTYPENEW, main = "Events with Highest Property Damage", ylab = "Amount of Damage ($ billions in Dec 2011 $ terms)", col = "red")
barplot(cropDMGtop$total_equivdamage/(10^9), las = 3, names.arg = cropDMGtop$EVTYPENEW, main = "Events with Highest Crop Damage", ylab = "Amount of Damage ($ billions in Dec 2011 $ terms)", col = "yellow")
```

**Figure 2: Events with Highest Property and crop Damage in Billions**
NB: figures have been asjusted to December 2011 dollars using the total Ip.50001.S index

### Trends on major public health and economy

The plots below show the annual fatalities for Tornados (black line) and Heat (red line; both Excessive Heat and Heat added together), and the annual property + crop damage for tornados (black line) and flood (red line).

``` {r AnnualTrend,fig.width=9,fig.height=3}
tornado_and_heat <- stormdata[grep("TORNADO|HEAT",stormdata$EVTYPENEW),]
tornado_and_heat$tornado <- ifelse(tornado_and_heat$EVTYPE=="TORNADO",TRUE,FALSE)
fatality_year <- data.frame(tornado=tornado_and_heat$tornado,year=year(tornado_and_heat$BGN_DATE),fatalities = tornado_and_heat$FATALITIES)
fatality_year <- fatality_year %>% group_by(tornado,year) %>% summarise(tot = sum(fatalities), count = n())
ymax <- max(fatality_year$tot)
propDMGyear <- propDMG %>% group_by(EVTYPENEW,year) %>% summarise(total_damage = sum(TOTDMG), total_equivdamage = sum(EQUIVTOTDMG), count = n())
cropDMGyear <- cropDMG %>% group_by(EVTYPENEW,year) %>% summarise(total_damage = sum(TOTDMG), total_equivdamage = sum(EQUIVTOTDMG), count = n())
# events with highest property damage
propDMGflood <- propDMGyear[propDMGyear$EVTYPENEW=="FLOOD",]
propDMGtornado <- propDMGyear[propDMGyear$EVTYPENEW=="TORNADO",]
cropDMGflood <- cropDMGyear[cropDMGyear$EVTYPENEW=="FLOOD",]
cropDMGtornado <- cropDMGyear[cropDMGyear$EVTYPENEW=="TORNADO",]
totDMGflood <- data.frame(year=propDMGflood$year,tot=propDMGflood$total_equivdamage + cropDMGflood$total_equivdamage)
totDMGtornado <- data.frame(year=propDMGtornado$year,tot=propDMGtornado$total_equivdamage)
totDMGtornado$tot[(62-18):62] <- totDMGtornado$tot[(62-18):62] + cropDMGtornado$total_equivdamage
ymaxDMG <- max(c(totDMGflood$tot,totDMGtornado$tot))
# events with highest injuriess
par(mfrow = c(1, 2), mar = c(4, 4, 4, 4), mgp = c(3, 1, 0), cex = 0.8)
plot(x=fatality_year[fatality_year$tornado==TRUE,]$year,y=fatality_year[fatality_year$tornado==TRUE,]$tot, type = "l", main = "Fatalities per year", xlab = "year", ylab = "Fatalities", col = 1, ylim = c(0,ymax))
points(x=fatality_year[fatality_year$tornado==FALSE,]$year,y=fatality_year[fatality_year$tornado==FALSE,]$tot, type = "l", col = 2)
legend(x="topleft",legend=c("TORNADO","HEAT"),lty=1,col = c(1,2))
plot(x=totDMGtornado$year,y=totDMGtornado$tot/(10^9), type = "l", main = "Property and Crop Damage per year", xlab ="year", ylab = "Amount of Damage ($ billions in Dec 2011 $ terms)", col = 1, ylim = c(0,ymaxDMG/(10^9)))
points(x=totDMGflood$year,y=totDMGflood$tot/(10^9), type = "l", col = 2)
legend(x="topleft",legend=c("TORNADO","FLOOD"),lty=1,col = c(1,2))
```

**Figure 3: Fatalities and Damage per year for the two most severe events of each**

## Conclusion

The graphs show that in terms of public health Tornados are significantly more impactful than most other sources.  So it suggests further research into how to reduce the health risk, such as warning systems, safety standards and shelters.  The effect of Heat on public health is also significant.

Flood had a major economic impact on both property and crops. So it suggests further research into how to reduce the economic impact of flood.  Further inquiry could be into the location of flooding and how damage could be further mitigated.  Possible new building standards or regulations in flood prone areas.  Flash flood and flood also feature in the top health risks.

Tornado, Flood (including Flash flood) and Excessive Heat (including Heat) appear to be the events that most impact on public health and the economy.

The annual totals do not show any trends that suggest any investment made to date has impacted the public health risk or economic impact from severe weather events in the two most significant contributors. Heat and Flood both have some outliers that could warrant further investigation.